Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data

### 2023-02-14 covid_nrs_place_of_death_data_2023.Rmd
cleaned up the processing of cause of death data; found discrepancy between 5-year weekly averages for 2020 and 2021 CoD data, which supposedly should be the same?

# TODO: this will eventually only have 2023 data, change when that happens
weekly_deaths_cause_of_death_2022_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

@@

weekly_deaths_cause_of_death_2021 <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data_2021,
    .f = ~readxl::read_excel(files_historical_deaths$`2021`, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

In the `2015-2019` period there were no Covid-19 deaths, so the historical average should be `NA`, which can possibly be interpreted as 0 for the purposes of comparing a year to the historical average.
For the `2016-2019, 2021` period the average number of covid deaths should be equivalent to the deaths in 2021 - we wouldn't want to make an average between 4 years of 0 deaths and one year with any deaths! Similarly if the period were `2017-2019, 2021-2022`, we would compute the average for 2021-2022, and so on. The year 2020 is considered exceptional and so is not used for computing the historical averages in 2021 and onwards.

```{r}
# TODO: change this so it works for 2023 and onwards

## this is an ugly hack - we set average weekly deaths due to C19 to NA for 2021
## and we set the average covid deaths for 2015-2019, 2021 to the same as in 2021
## however, for the latter, we set the dates to 2022
average_weekly_covid_deaths_2021_onwards <-
  bind_rows(
    ## set 2021 averages to NA
    weekly_deaths_cause_of_death_2021 %>%
    filter(cause_of_death == "COVID-19") %>%
    mutate(number_of_deaths = NA_real_) %>%
    rename(deaths_mean = number_of_deaths),
    ## use 2021 weekly deaths as 2022 historic average
    weekly_deaths_cause_of_death_2021 %>%
    filter(cause_of_death == "COVID-19") %>%
    mutate(year = 2022) %>%
    select(-c(week_number_run_over, date_w_c)) %>%
    rename(deaths_mean = number_of_deaths) %>%
    inner_join(weekly_deaths_cause_of_death_2022_onwards %>% 
                select(year, week_number, week_number_run_over, date_w_c, place_of_death, cause_of_death)
              )
  ) %>%
  mutate(historic_period = historic_period_depending_on_year(year))

average_weekly_noncovid_deaths_2021_onwards <-
    ## extract 5-year averages from most recent file for 2021 onwards
    ## this doesn't include covid, which we've manually extracted above!
  bind_rows(
    ## 2022+ data
    map_dfr(
      .x = cell_ranges_for_causes_of_death_data,
      .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
      .id = "place_of_death"
    ),
    ## 2021 data
    map_dfr(
      .x = cell_ranges_for_causes_of_death_data_2021,
      .f = ~readxl::read_excel(files_historical_deaths$`2021`, sheet = sheet_cause_of_death_data, range = .x), 
      .id = "place_of_death"
    )
  ) %>%
    janitor::clean_names() %>%
    rename(year = registration_year, date_w_c = week_beginning) %>%
    select(place_of_death, year, week_number, date_w_c, matches("five_year_average")) %>%
    pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "deaths_mean") %>%
    mutate(
      year = as.integer(year),
      cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
      cause_of_death = return_pretty_cause_of_death(cause_of_death),
      cause_of_death = if_else(cause_of_death == "All causes", "All", cause_of_death),
      date_w_c = as.Date(date_w_c, origin = "1899-12-30"),
      week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
      historic_period = historic_period_depending_on_year(year)
    ) %>%
    select(-name) %>%
    drop_na(deaths_mean)  # remove blank entries - these are for the current year but for dates in the future

historical_average_cause_of_death_2021_onwards <-
  bind_rows(
    ## the hacked averages for 2021 and 2022 above
    average_weekly_covid_deaths_2021_onwards,
    ## the non-covid deaths in the 'main report' files
    average_weekly_noncovid_deaths_2021_onwards
  )
```

## Merge 2020 & 2021 data with current data

Note: the 5-year average for 2022 (spanning 2016-2019 & 2021) is only published along with the running weekly counts of deaths by cause of death; it's just that NRS has decided to only publish them at the same time as new data comes out! This is a bit of a bummer - it means we can't populate the whole year of 'historical context' data for 2022.

They could be calculated by hand if the 2015-2019 weekly data were available broken down by cause of death - as it stands, only the average 2015-2019 figure is available.

```{r}
weekly_deaths_cause_of_death <-
  bind_rows(
    weekly_deaths_cause_of_death_2022_onwards,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-deaths_mean) %>%
      filter(cause_of_death != "All")  # don't need all cause data in causes breakdown!
  )

historical_average_cause_of_death <-
  bind_rows(
    historical_average_cause_of_death_2021_onwards,
    weekly_deaths_cause_of_death_2020 %>% 
      select(-number_of_deaths) %>% 
      mutate(historic_period = historic_period_depending_on_year(year))
  )
```

## Historical annual average by cause

Cause of death data for 2015-2019 was only given in terms of weekly averages over 5 years. With the later historic periods (2016-2019 + 2021 and so on) I compute a weighted average between the 5-year average and the current weekly data.

The below compromise figure is no longer used, but the reasoning for the weighted average is the same (except for the quirk of the denominator of 6 in the below case, where I'm including the 5 years of historic data plus the one year of current data, hence 6 years in the denominator):

I can approximate this by taking the sum of the average weekly deaths for `2015-2019`; for the `2016-2019 + 2021` period this isn't readily available because NRS only publish the weekly averages alongside the weekly updated data, but not for the entire year. Therefore I'll just calculate the `2015-2019` period average for now, and a compromise `2015-2019 + 2021` average:

$$\frac{5}{6}average\_deaths_{2015-2019} + \frac{1}{6}deaths_{2021}$$

```{r}
# TODO: incorporate the 2016-2019 + 2021 period
historical_annual_average_cause_of_death_2015_2019 <-
  historical_average_cause_of_death %>%
  filter(historic_period == "2015-2019" & year == 2021) %>%  # this is slightly strange but some of the averages disagree, probably because the 2015-2019 data were updated in 2020, and the 2021 data were published with modified figures. NRS deaths data are truly a living document!
  group_by(place_of_death, cause_of_death, historic_period) %>%
  summarise(
    annual_deaths_mean = sum(deaths_mean),
    .groups = "drop"
  )

deaths_by_cause_2021 <-
  weekly_deaths_cause_of_death_2021 %>% 
  group_by(place_of_death, cause_of_death) %>% 
  summarise(deaths_2021 = sum(number_of_deaths), .groups = "drop")

historical_annual_average_cause_of_death <-
  bind_rows(
    historical_annual_average_cause_of_death_2015_2019
    ,
    ## compromise figures for 2015-2019 + 2021
    historical_annual_average_cause_of_death_2015_2019 %>%
    filter(cause_of_death != "All") %>%
    left_join(
      deaths_by_cause_2021, by = c("place_of_death", "cause_of_death")
    ) %>%
    mutate(
      annual_deaths_mean = (5/6 * annual_deaths_mean) + (1/6 * deaths_2021),
      deaths_2021 = NULL,
      historic_period = "2015-2019, 2021"
    )
  )
```

>>>>

# TODO: this will eventually only have 2023 data, change when that happens
weekly_deaths_cause_of_death_2022_onwards <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data,
    .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = number_of_deaths)
  # filter(!is.na(number_of_deaths))   # this will also remove the historic mean values, so I've opted not to remove blank entries!
```

@@

~~Note that due to how I originally processed these files, the weekly CoD data for 2020 *includes* the 5-year historic weekly average for 2015-2019, whereas I calculate that manually later for the 2021 data and later.~~ This is now changed so that the weekly CoD data is now

@@

weekly_deaths_cause_of_death_2021 <-
  map_dfr(
    .x = cell_ranges_for_causes_of_death_data_2021,
    .f = ~readxl::read_excel(files_historical_deaths$`2021`, sheet = sheet_cause_of_death_data, range = .x), 
    .id = "place_of_death"
  ) %>%
  janitor::clean_names() %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  select(-matches("_excess$")) %>%  # remove excess deaths, not currently used in analysis
  # select(-matches("_excess$|_average$")) %>%  # we can compute excess & averages later
  pivot_longer(cols = 5:ncol(.), names_to = "name", values_to = "number_of_deaths") %>%
  filter(!str_detect(name, "all_causes")) %>%  # we can calculate that from the others
  mutate(
    year = as.integer(year),
    measure_type = if_else(condition = str_detect(name, "_five_year_average"), true = "deaths_mean", false = "number_of_deaths"),
    cause_of_death = str_remove_all(name, "_five_year_average|_deaths"),
    cause_of_death = return_pretty_cause_of_death(cause_of_death),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020)  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
  ) %>%
  select(-name) %>%
  pivot_wider(names_from = measure_type, values_from = number_of_deaths) %>%
  filter(!is.na(number_of_deaths))  # remove blank entries - these are for the current year but for dates in the future
```

@@

Cause of death data comes on a per-year basis, with the current year's data including the preceding year; at some point mid-year, this changes, and only the current year's data are included, and the preceding year's data need to be loaded from the last release that included them. (e.g. mid-2022, causes of death for 2021 stopped being released alongside 2022 data, so I load them from the last 2022 release that included 2021 causes of death also). They therefore come as one dataset per year, with the current year as a separate dataset. The weekly figures for cause of death are included alongside the 5-year historic average for that week except for Covid, which has been set to NA for all data up and including 2023 data as of February 2023.

In the `2015-2019` period there were no Covid-19 deaths, so the historical average should be `NA`, which can possibly be interpreted as 0 for the purposes of comparing a year to the historical average.

For the `2016-2019, 2021` period the average number of covid deaths should be equivalent to the deaths in 2021 - we wouldn't want to make an average between 4 years of 0 deaths and one year with any deaths! Similarly if the period were `2017-2019, 2021-2022`, we would compute the average for 2021-2022, and so on. The year 2020 is considered exceptional and so is not used for computing the historical averages in 2021 and onwards.

```{r}
# TODO: include 2023 data once it's reported separately

# TODO: I still need to separately process the most recent year because the average values aren't provided

## here I separately merge the datasets for cause of death data from different years
## I treat covid deaths separately to recalculate a realistic esstimate of average deaths in prior years
weekly_deaths_cause_of_death_noncovid <-
  bind_rows(
    weekly_deaths_cause_of_death_2020,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2022_onwards
  ) %>%
  filter(cause_of_death != "COVID-19")

weekly_deaths_cause_of_death_covid <-
  bind_rows(
    weekly_deaths_cause_of_death_2020,
    weekly_deaths_cause_of_death_2021,
    weekly_deaths_cause_of_death_2022_onwards
  ) %>%
  filter(cause_of_death == "COVID-19") %>%
  group_by(year) %>%
  group_map(
    .f = function(data_tbl, year_tbl) {
      current_year <- year_tbl$year  # get year from the function arguments
      year_range <- infer_year_range_for_calculating_covid_deaths(current_year)  # work out which years of data to use
      weekly_means <-  
        weekly_deaths_cause_of_death_covid %>%
        filter(year %in% year_range) %>%  # filter the covid data to the relevant years to compute from
        create_week_53_data_for_years_with_52_weeks() %>%  # create 53rd week if needed
        group_by(place_of_death, cause_of_death, week_number) %>%
        summarise(
          deaths_mean = mean(number_of_deaths, na.rm=TRUE),
          # year_range = paste0(year_range,collapse=","),  # for testing
          .groups = "drop"
        )
      data_tbl %>%
        mutate(year = current_year) %>%  # this gets removed in the grouping, so I re-add it here
        select(-deaths_mean) %>%  # remove the NA/0 mean deaths
        create_week_53_data_for_years_with_52_weeks() %>%  # create 53rd week if needed
        left_join(weekly_means, by = c("place_of_death", "cause_of_death", "week_number"))  # merge the mean deaths
    }
  ) %>%
  bind_rows()  # merge into single tibble from list

## re-merge non-covid and covid deaths again
# TODO: make CoD, PoD factors
weekly_deaths_cause_of_death <-
  bind_rows(
    weekly_deaths_cause_of_death_noncovid,
    weekly_deaths_cause_of_death_covid
  ) %>%
  mutate(historic_period = historic_period_depending_on_year(year))

## note that these might differ from actual historic figures a bit because of
## my hack to create a fake 53rd week in years with only 52 weeks for easier 
## comparison

# TODO: I get different numbers of 5-year annual average deaths between 2020 and 2021, how come?
historical_annual_average_cause_of_death <-
  weekly_deaths_cause_of_death %>%
  group_by(year, historic_period, place_of_death, cause_of_death) %>%
  summarise(
    annual_deaths_mean = sum(deaths_mean),
    .groups = "drop"
  ) %>% 
  filter(year!=2020)  # 2020 and 2021 should have identical annual means so I remove one %>%
  select(-year)

## checking discrepancy:
## I get different numbers of 5-year annual average deaths between 2020 and 2021, how come?
weekly_deaths_cause_of_death %>% filter(year %in% 2020:2021) %>% select(place_of_death, cause_of_death, year, week_number, deaths_mean) %>%
  pivot_wider(names_from = year, values_from = deaths_mean) %>%
  filter(`2020`!=`2021`) %>%
  arrange(desc(week_number)) %>%
  summarise(across(matches("\\d{4}"), ~sum(.x, na.rm=TRUE)))
## looks like 2020 and 2021 don't actually use the same 5-year weekly average deaths by cause!!!
# TODO: act on this knowledge...
```
